Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-03-28 _tex/paper2.tex
Built site for gh-pages

Notably, every SCM has an associated DAG
\citep{Pearl_et_al_2016, Cinelli_et_al_2020}. A DAG is a \emph{graph}
consisting of nodes connected by edges, where nodes represent random
variables. The term \emph{directed} indicates that edges or arrows
extend from one node to another, indicating the direction of causal
influence. The absence of an edge implies no direct relationship between
the nodes. The term \emph{acyclic} means that the causal influences do
not form loops, ensuring the influences do not cycle back on themselves
\citep{McElreath_2020}. DAGs conventionally depict observed variables as
solid black circles and unobserved (latent) variables as open circles
\citep{Morgan_et_al_2014}. They also display conditioned variables by
surrounding them with a square. Here, \emph{conditioning} refers to
restricting the statistical analysis to a subset of the population
defined by the values of the given variable \citep[pp.~32]{Neal_2020}.
Although DAGs often omit exogenous variables for simplicity, the DAGs
presented in this section includes exogenous variables to improve
clarity and reveal potential issues related to conditioning and
confounding \citep{Cinelli_et_al_2020}.

@@

\[
I = \begin{bmatrix}
1 \\
2 \\
3 \\
4 \\
5 \\
\end{bmatrix} ; \;
J = \begin{bmatrix}
1 \\
2 \\
3 \\
\end{bmatrix} ; \;
IA = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
2 & 1 \\
2 & 2 \\
3 & 1 \\
3 & 2 \\
4 & 1 \\
4 & 2 \\
5 & 1 \\
5 & 2 \\
\end{bmatrix} ; \;
JK = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
1 & 3 \\
2 & 1 \\
2 & 2 \\
2 & 3 \\
3 & 1 \\
3 & 2 \\
3 & 3 \\
\end{bmatrix} ; \;
R = \begin{bmatrix}
1 & 1 & 1 & 2 & 1 & 1 \\
1 & 1 & 1 & 2 & 1 & 2 \\
1 & 1 & 1 & 2 & 1 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 5 & 1 & 1 & 1 \\
1 & 1 & 5 & 1 & 1 & 2 \\
1 & 1 & 5 & 1 & 1 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 5 & 1 & 3 & 1 \\
1 & 1 & 5 & 1 & 3 & 2 \\
1 & 1 & 5 & 1 & 3 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
5 & 1 & 5 & 2 & 3 & 1 \\
5 & 1 & 5 & 2 & 3 & 2 \\
5 & 1 & 5 & 2 & 3 & 3 \\
\end{bmatrix}
\]

@@

The conceptual-population model integrates two \emph{hierarchical
structural components} to examine how different teaching methods
influence students' writing ability. This approach builds on the
principles of Structural Equation Modeling (SEM)
\citep[pp.~138]{Hoyle_et_al_2023} and Item Response Theory (IRT)
\citetext{\citealp[chap.~6]{Fox_2010}; \citealp[chap.~24]{vanderLinden_et_al_2017_I}}.
Each structural component defines how observed or latent variables
affect the primary latent variable of interest
\citep{Everitt_et_al_2010}. Their hierarchical nature enables
researchers to test hypotheses while accounting for uncertainties
associated with trait estimation (see Section~\ref{sec-theory-issue2}),
particularly when stimuli exhibit a hierarchical structure (see
Section~\ref{sec-theory-issue1b}).

The top branch of DAG~\ref{fig-cj09_dag} depicts the first component.
This component describes how the students' writing-quality trait
\((T_{I})\) and a set of \emph{relevant} text-related variables
\((X_{IA})\) (e.g., text length) causally influence the texts'
written-quality trait \((T_{IA})\). The DAG also highlights the
idiosyncratic errors \((e_{IA})\), which capture variations in the
texts' trait that neither the students' trait nor other relevant
text-related variables can explain. In this structure, \(T_{I}\) and
\(e_{IA}\) are latent vectors of length \(n_{I}\) and
\((n_{I} \cdot n_{A})\), respectively. Additionally, \(X_{IA}\) is an
observed matrix with a dimension of \((n_{I} \cdot n_{A})\) rows and
\(q_{IA}\) columns (variables).

Additionally, the DAG's top branch illustrates how a set of
\emph{relevant} student-related variables \((X_{I})\), such as the
teaching method, causally influence the students' writing-quality trait
\((T_{I})\). The DAG also includes the idiosyncratic errors \((e_{I})\),
which capture variations in the students' traits that the relevant
student-related variables cannot explain. Here, \(e_{I}\) is a latent
vector of length \(n_{I}\), and \(X_{I}\) can be an observed matrix with
\(n_{I}\) rows and \(q_{I}\) columns (variables).

Here, \emph{relevant} refers to variables that satisfy the
\emph{backdoor criterion} \citep[pp 37]{Neal_2020}, that is, they belong
to a \emph{sufficient adjustment set}
\citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}. A
\emph{sufficient} set (potentially empty) blocks all non-causal paths
between a predictor and an outcome without opening new ones
\citep{Pearl_2009}.

@@

Similarly, the bottom branch of DAG~\ref{fig-cj09_dag} depicts the
second component. This component describes how the judges' bias
\((B_{J})\) and a set of \emph{relevant} judgment-related variables
\((Z_{JK})\) (e.g., the number of judgments a judge makes) causally
influence the biases associated with each text \(B_{JK}\). The DAG also
highlights the idiosyncratic errors \((e_{JK})\), which capture
variations in the texts' bias that neither the judges' bias nor other
relevant judgment-related variables can explain. In this description,
\(B_{J}\) and \(e_{JK}\) are latent vectors of length \(n_{J}\) and
\((n_{J} \cdot n_{K})\), respectively. Additionally, \(Z_{JK}\) is an
observed matrix with a dimension of \((n_{J} \cdot n_{K})\) rows and
\(q_{JK}\) columns.

Moreover, the DAG's bottom branch describes how a set of \emph{relevant}
judge-related variables \((Z_{J})\), such as judgment expertise,
causally influence the judges' bias \((B_{J})\). The DAG also accounts
for idiosyncratic errors \((e_{J})\), capturing the variations in the
judges' bias that relevant judge-related variables cannot explain. Here,
\(e_{J}\) is a latent vector of length \(n_{J}\), and \(Z_{J}\) is an
observed matrix with \(n_{J}\) rows and \(q_{J}\) columns.

@@

To facilitate the interpretation of the sample-comparison model, we
first define the data sample using several binary vector variables:
\(S_{I}\), \(S_{J}\), \(S_{IA}\), \(S_{JK}\), and \(S\) as follows.

\begin{equation}\phantomsection\label{eq-mat21}{
S_{I} = \begin{bmatrix}
i_{(1)} \\
i_{(2)} \\
\vdots \\
i_{(h)} \\
\vdots \\
i_{(i)} \\
\vdots \\
i_{(nI)}
\end{bmatrix} ; \;
S_{J} = \begin{bmatrix}
j_{(1)} \\
j_{(2)} \\
\vdots \\
j_{(j)} \\
\vdots \\
j_{(nJ)}
\end{bmatrix} ; \;
S_{IA} = \begin{bmatrix}
ia_{(1)} \\
ia_{(2)} \\
\vdots \\
ia_{(ia)} \\
\vdots \\
ia_{(nI \cdot nA)} \\
\end{bmatrix} ; \;
S_{JK} = \begin{bmatrix}
jk_{(1)} \\
jk_{(2)} \\
\vdots \\
jk_{(jk)} \\
\vdots \\
jk_{(nJ \cdot nK)} \\
\end{bmatrix}; \;
S = \begin{bmatrix}
s_{(1)} \\
s_{(2)} \\
\vdots \\
s_{(s)} \\
\vdots \\
s_{(n)} \\
\end{bmatrix}
}\end{equation}

Each element of the vectors \(S_{I}\) and \(S_{J}\) is a binary value
that indicates the presence \((1)\) or absence \((0)\) of data in the
\(I\) and \(J\) vectors, as defined in equation (\ref{eq-mat11}).
Therefore, the vector \(S_{I}\) contains \(n_{I}\) elements, and the
vector \(S_{J}\) contains \(n_{J}\) elements. For example, we define
each element of the vector \(S_{I}\) as in equation (\ref{eq-mat22}),
and we apply a similar definition to \(S_{J}\).

@@

Each element of the vectors \(S_{IA}\) and \(S_{JK}\) is defined like
\(i_{(i)}\) with vector \(I\), but applied to data in the matrices
\(IA\) and \(JK\), as defined in equation (\ref{eq-mat11}). Therefore,
the vector \(S_{IA}\) contains \(n_{I} \cdot n_{A}\) elements, and the
vector \(S_{JK}\) contains \(n_{J} \cdot n_{K}\) elements. For example,
we define each element of the vector \(S_{IA}\) as in equation
(\ref{eq-mat23}), and we apply a similar definition to \(S_{JK}\).

@@

The equation shows that the observed values in the \(S_{I}\) and
\(S_{J}\) vectors determine the binary values in the \(S_{IA}\) and
\(S_{JK}\) vectors, through the \(IA\) and \(JK\) matrices,
respectively. Specifically, researchers cannot observe stimuli and
judgments for individuals and judges they did not sample initially.

Finally, the vector \(S\) in equation (\ref{eq-mat21}) contains
\(n = \binom{n_{I} \cdot n_{A}}{2} \cdot n_{J} \cdot n_{K}\) elements,
with each element defined in equation (\ref{eq-mat24}). This equation
shows that the observed values in \(S_{IA}\) and \(S_{JK}\) determine
the binary values in \(S\) through the \(R\) matrix. Specifically,
researchers can only observe a subset of judges' comparisons for a
subset of stimuli generated by a subset of individuals.

\begin{equation}\phantomsection\label{eq-mat24}{
s_{(s)} = \begin{cases} 
1 & \text{if the data for the } i,a \text{ with the } h,b \text{ and the } j,k \text{ elements of } R \text{ are observed} \\
0 & \text{if the data for the } i,a \text{ with the } h,b \text{ and the } j,k \text{ elements of } R \text{ are missing}
\end{cases}
}\end{equation}

With the definition of \(S\), we can now integrate the sample mechanism
into the conceptual-population model. Figure~\ref{fig-cj12} illustrates
this model. Following the convention of
\citet[pp.~499-516]{McElreath_2020} and
\citet[pp.~6]{Deffner_et_al_2022}, the DAG~\ref{fig-cj12_dag} depicts
the conceptual-population outcome \(O^{cp}_{R}\) as an unobserved
variable, emphasizing the researchers' inability to access such an
outcome due to the sampling mechanism. Additionally, the DAG depicts the
conditioned \emph{sample design} variable \(S\) causally influencing the
sample-comparison outcome \(O^{sc}_{R}\). This variable shows that the
sample-comparison outcome is a subset drawn from the
conceptual-population outcome \(O^{cp}_{R}\).

@@

\[
\begin{aligned}
  O_{R} & = O^{sc}_{R} \odot C \\
  O^{sc}_{R} & = O^{cp}_{R} \odot S \\
  O^{cp}_{R} & \overset{iid}{\sim} \text{Bernoulli} \left[ \text{inv\_logit}( D_{R} ) \right] \\
  D_{R} & = \left( T_{IA}[R_{1},P_{1}] - T_{IA}[R_{2},P_{2}] \right) + B_{JK}[U,V] \\
  T_{IA} & = T_{I} + \beta_{XA} X_{IA} + e_{IA} \\
  T_{I} & = \beta_{XI} X_{I} + e_{I} \\
  B_{JK} & = B_{J} + \beta_{ZK} Z_{JK} + e_{JK} \\
  B_{J} & = \beta_{ZJ} Z_{J} + e_{J} \\
  e_{I} & \overset{iid}{\sim} \text{Normal}(0,1) \\
  e_{J} & \overset{iid}{\sim} \text{Normal}(0,1) \\
  e_{IA} & \overset{iid}{\sim} \text{Normal}(0, p_{IA}) \\  
  e_{JK} & \overset{iid}{\sim} \text{Normal}(0, p_{JK})
\end{aligned}
\]

\[
\begin{aligned}
  O_{R} & := f_{C}(O^{sc}_{R}, C) \\
  O^{sc}_{R} & := f_{S}(O^{cp}_{R}, S) \\
  O^{cp}_{R} & := f_{O}(D_{R}) \\
  D_{R} & := f_{D}(T_{IA}, B_{JK}) \\
  T_{IA} & := f_{T}(T_{I}, X_{A}, e_{IA}) \\
  T_{I} & := f_{T}(X_{I}, e_{I}) \\
  B_{JK} & := f_{B}(B_{J}, Z_{K}, e_{JK}) \\
  B_{J} & := f_{B}(Z_{J}, e_{J}) \\
  e_{I} & \:\bot\:\{ e_{J}, e_{IA}, e_{JK} \} \\
  e_{J} & \:\bot\:\{ e_{IA}, e_{JK} \} \\
  e_{IA} & \:\bot\:e_{JK}
\end{aligned}
\]

\caption{\label{fig-cj15}Sample-comparison model, including sample and
comparison mechanisms and assuming equal discriminal dispersions for the
student's traits}

@@

\subcaption{\label{fig-cj16_stat}Statistical model}

@@

\subcaption{\label{fig-cj16_scm}SCM}

@@

\caption{\label{fig-cj16}Sample-comparison model, including sample and
comparison mechanisms and assuming different discriminal dispersions for
the student's traits}

>>>>

Notably, every SCM has an associated DAG
\citep{Pearl_et_al_2016, Cinelli_et_al_2020}. A DAG is a \emph{graph}
consisting of nodes connected by edges, where nodes represent random
variables. The term \emph{directed} indicates that edges or arrows
extend from one node to another, indicating the direction of causal
influence. The absence of an edge implies no direct relationship between
the nodes. The term \emph{acyclic} means that the causal influences do
not form loops, ensuring the influences do not cycle back on themselves
\citep{McElreath_2020}. DAGs conventionally depict observed variables as
solid black circles and unobserved (latent) variables as open circles
\citep{Morgan_et_al_2014}. Although DAGs often omit exogenous variables
for simplicity, the DAGs presented in this section includes exogenous
variables to improve clarity and reveal potential issues related to
conditioning and confounding \citep{Cinelli_et_al_2020}.

@@

\begin{equation}\phantomsection\label{eq-mat13}{
I = \begin{bmatrix}
1 \\
2 \\
3 \\
4 \\
5 \\
\end{bmatrix} ; \;
J = \begin{bmatrix}
1 \\
2 \\
3 \\
\end{bmatrix} ; \;
IA = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
2 & 1 \\
2 & 2 \\
3 & 1 \\
3 & 2 \\
4 & 1 \\
4 & 2 \\
5 & 1 \\
5 & 2 \\
\end{bmatrix} ; \;
JK = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
1 & 3 \\
2 & 1 \\
2 & 2 \\
2 & 3 \\
3 & 1 \\
3 & 2 \\
3 & 3 \\
\end{bmatrix} ; \;
R = \begin{bmatrix}
1 & 1 & 1 & 2 & 1 & 1 \\
1 & 1 & 1 & 2 & 1 & 2 \\
1 & 1 & 1 & 2 & 1 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 5 & 1 & 1 & 1 \\
1 & 1 & 5 & 1 & 1 & 2 \\
1 & 1 & 5 & 1 & 1 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
1 & 1 & 5 & 1 & 3 & 1 \\
1 & 1 & 5 & 1 & 3 & 2 \\
1 & 1 & 5 & 1 & 3 & 3 \\
\vdots & \vdots & \vdots & \vdots & \vdots & \vdots\\
5 & 1 & 5 & 2 & 3 & 1 \\
5 & 1 & 5 & 2 & 3 & 2 \\
5 & 1 & 5 & 2 & 3 & 3 \\
\end{bmatrix}
}\end{equation}

@@

Building on the principles of Structural Equation Modeling (SEM)
\citep[pp.~138]{Hoyle_et_al_2023} and Item Response Theory (IRT)
\citetext{\citealp[chap.~6]{Fox_2010}; \citealp[chap.~24]{vanderLinden_et_al_2017_I}},
the conceptual-population model integrates two \emph{hierarchical
structural components} to examine how different teaching methods
influence students' writing ability. Each structural component defines
how observed or latent variables affect the primary latent variable of
interest \citep{Everitt_et_al_2010}. Their hierarchical nature enables
researchers to test hypotheses while accounting for uncertainties
associated with trait estimation (see Section~\ref{sec-theory-issue2}),
particularly when stimuli exhibit a hierarchical structure (see
Section~\ref{sec-theory-issue1b}).

The top branch of DAG~\ref{fig-cj09_dag} illustrates the first
component, where \emph{relevant} \footnote{Here, \emph{relevant} refers
  to variables that satisfy the \emph{backdoor criterion} \citep[pp
  37]{Neal_2020}, that is, they belong to a \emph{sufficient adjustment
  set} \citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}. A
  \emph{sufficient} set (potentially empty) blocks all non-causal paths
  between a predictor and an outcome without opening new ones
  \citep{Pearl_2009}. While these topics are beyond the scope of this
  study, readers seeking a deeper understanding can refer to
  introductory papers such as \citet{Pearl_2010}, \citet{Rohrer_2018},
  \citet{Pearl_2019}, and \citet{Cinelli_et_al_2020}, as well as
  introductory books like \citet{Pearl_et_al_2018}, \citet{Neal_2020}
  and \citet{McElreath_2020} useful. For more advanced study, seminal
  papers such as \citet{Neyman_et_al_1923}, \citet{Rubin_1974},
  \citet{Spirtes_et_al_1991}, and \citet{Sekhon_2009}, along with books
  such as \citet{Pearl_2009}, \citet{Morgan_et_al_2014} and
  \citet{Hernan_et_al_2020}, are recommended.} student-related variables
\((X_{I})\), such as the teaching method, causally influence the latent
variable representing students' writing-quality trait \((T_{I})\). It
also shows how \(T_{I}\) and \emph{relevant} \footnote{refer to footnote
  1.} text-related variables \((X_{IA})\) (e.g., text length) causally
influence the texts' written-quality trait \((T_{IA})\), the primary
latent variable of interest. Additionally, the branch accounts for
idiosyncratic errors: \(e_{I}\) captures unexplained variations in
students' traits unexplained by \(X_{I}\), while \(e_{IA}\) captures
variations in texts' traits unexplained by \(T_{I}\) or \(X_{IA}\).
Here, \(X_{I}\) is an observed matrix with \(n_{I}\) rows and \(q_{I}\)
columns (variables), and both \(e_{I}\) and \(T_{I}\) are latent vectors
of length \(n_{I}\). Similarly, \(X_{IA}\) is an observed matrix with
dimension \((n_{I} \cdot n_{A})\) rows and \(q_{IA}\) columns
(variables), while \(e_{IA}\) and \(T_{IA}\) are latent vectors of
length \(n_{I} \cdot n_{A}\).

@@

Similarly, the bottom branch of DAG~\ref{fig-cj09_dag} depicts the
second component, where \emph{relevant} \footnote{refer to footnote 1.}
judge-related variables \((Z_{J})\), such as judgment expertise,
causally influence the latent variable representing judges' bias
\((B_{J})\). It also shows how \(B_{J}\) and \emph{relevant} \footnote{refer
  to footnote 1.} judgment-related variables \((Z_{JK})\) (e.g., the
number of judgments a judge makes) causally influence the biases
associated with each text \(B_{JK}\), the primary latent variable of
interest. Additionally, the branch accounts for idiosyncratic errors:
\(e_{J}\) captures unexplained variations in judges' bias unexplained by
\(Z_{J}\), while \(e_{JK}\) captures variations in judgments unexplained
by \(B_{J}\) or \(Z_{JK}\). Here, \(Z_{J}\) is an observed matrix with
\(n_{J}\) rows and \(q_{J}\) columns (variables), and both \(e_{J}\) and
\(B_{J}\) are latent vectors of length \(n_{J}\). Similarly, \(Z_{JK}\)
is an observed matrix with dimension \((n_{J} \cdot n_{K})\) rows and
\(q_{JK}\) columns (variables), while \(e_{JK}\) and \(B_{JK}\) are
latent vectors of length \(n_{J} \cdot n_{K}\).

@@

To incorporate the sampling mechanism and facilitate the interpretation
of the sample-comparison model, we first define the data sampling
process using the binary vector variables \(S_{I}\), \(S_{J}\),
\(S_{IA}\), and \(S_{JK}\) as follows:

\begin{equation}\phantomsection\label{eq-mat21}{
S_{I} = \begin{bmatrix}
i_{(1)} \\
i_{(2)} \\
\vdots \\
i_{(h)} \\
\vdots \\
i_{(i)} \\
\vdots \\
i_{(nI)}
\end{bmatrix} ; \;
S_{J} = \begin{bmatrix}
j_{(1)} \\
j_{(2)} \\
\vdots \\
j_{(j)} \\
\vdots \\
j_{(nJ)}
\end{bmatrix} ; \;
S_{IA} = \begin{bmatrix}
ia_{(1)} \\
ia_{(2)} \\
\vdots \\
ia_{(ia)} \\
\vdots \\
ia_{(nI \cdot nA)} \\
\end{bmatrix} ; \;
S_{JK} = \begin{bmatrix}
jk_{(1)} \\
jk_{(2)} \\
\vdots \\
jk_{(jk)} \\
\vdots \\
jk_{(nJ \cdot nK)} \\
\end{bmatrix}
}\end{equation}

Where each element of the vectors \(S_{I}\) and \(S_{J}\) is a binary
value indicating the presence or absence of data in the \(I\) and \(J\)
vectors (see equation (\ref{eq-mat11})). Thus, the vectors \(S_{I}\) and
\(S_{J}\) contains \(n_{I}\) and \(n_{J}\) elements, respectively.
Specifically, we define the vector \(S_{I}\) as in equation
(\ref{eq-mat22}) and apply a similar definition to \(S_{J}\).

@@

Similarly, each element of the vectors \(S_{IA}\) and \(S_{JK}\) is a
binary value indicating the presence or absence of data in the matrices
\(IA\) and \(JK\) (see equation (\ref{eq-mat11})). Thus, the vectors
\(S_{IA}\) and \(S_{JK}\) contains \(n_{I} \cdot n_{A}\) and
\(n_{J} \cdot n_{K}\) elements, respectively. Specifically, we define
the vector \(S_{IA}\) as in equation (\ref{eq-mat23}) and apply a
similar definition to \(S_{JK}\).

@@

We can visualize these vectors more clearly using the example in
equation (\ref{eq-mat13}). Suppose the sample excludes the second
student and the third judge. Additionally, assume that the second text
from each individual and the third judgment from each judge are also
missing. Under these conditions, the vectors have the following
structure:

\begin{equation}\phantomsection\label{eq-mat24}{
S_{I} = \begin{bmatrix}
1 \\
0 \\
1 \\
1 \\
1
\end{bmatrix} ; \;
S_{J} = \begin{bmatrix}
1 \\
1 \\
0
\end{bmatrix} ; \;
S_{IA} = \begin{bmatrix}
1 \\
0 \\
0 \\
0 \\
1 \\
0 \\
1 \\
0 \\
1 \\
0 
\end{bmatrix} ; \;
S_{JK} = \begin{bmatrix}
1 \\
1 \\
0 \\
1 \\
1 \\
0 \\
0 \\
0 \\
0 
\end{bmatrix}
}\end{equation}

Equation (\ref{eq-mat24}) shows that missing values in the \(S_{I}\) and
\(S_{J}\) vectors directly determine those in the \(S_{IA}\) and
\(S_{JK}\) vectors. In other words, researchers can only observe texts
and judgments from students and judges initially included in the sample.
Additionally, it shows that the sum of observed elements in \(S_{I}\)
equals the sampled students \((n^{s}_{I})\) and that a similar sum in
vector \(S_{J}\) equals the sampled judges \((n^{s}_{J})\). Conversely,
the sum of observed elements in \(S_{IA}\) represents the total sampled
texts across all students \((n^{s}_{I} \cdot n^{s}_{A})\), while a
similar sum in vector \(S_{JK}\) represents the total sampled judgments
across all judges \((n^{s}_{J} \cdot n^{s}_{K})\).

Finally, we define the \emph{sample design} vector \(S\) in equation
(\ref{eq-mat25}), which maps each element of \(S_{IA}\) to every
component of \(S_{JK}\). Essentially, this vector serves as the sampled
equivalent of the \(R\) matrix defined in equation (\ref{eq-mat12}),
where the sum of its elements represents the total data sample
\(n^{s} = \binom{n^{s}_{I} \cdot n^{s}_{A}}{2} \cdot n^{s}_{J} \cdot n^{s}_{K}\).

\begin{equation}\phantomsection\label{eq-mat25}{
S = \begin{bmatrix}
s_{(1)} \\
s_{(2)} \\
\vdots \\
s_{(s)} \\
\vdots \\
s_{(n)} \\
\end{bmatrix}
}\end{equation}

With the definition of \(S\), we incorporate the sample mechanism into
the conceptual-population model. Following the convention of
\citet[pp.~499-516]{McElreath_2020} and
\citet[pp.~6]{Deffner_et_al_2022}, the DAG~\ref{fig-cj12_dag} represents
the conceptual-population outcome \(O^{cp}_{R}\) as unobserved,
emphasizing that researchers cannot directly access this outcome due to
the sampling mechanism.

Additionally, the DAG depicts the \emph{sample design} vector \(S\) as a
causal factor influencing the sample-comparison outcome \(O^{sc}_{R}\).
A square encloses this vector, indicating that it is a conditioned
variable. In this context, \emph{conditioning} means restricting the
statistical analysis to a specific subset of the population based on the
values of a given variable \citep[pp.~32]{Neal_2020}. Thus, the vector
\(S\) is responsible for selecting the subset of observation from the
conceptual-population outcome \(O^{cp}_{R}\), generating the
sample-comparison outcome \(O^{sc}_{R}\).

@@

when you talk about the variability, refer that the equal dispersion is
just assuming 1 for identification.

@@

@@

@@

\caption{\label{fig-cj15}Sample-comparison model, including sample and
comparison mechanisms and assuming different discriminal dispersions for
the student's traits}
